Autopolyploid establishment through polygenic adaptation

### 2024-06-05 doc/paper/paper.tex
up

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N, N_k$ & total population size, population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability that an unreduced gamete is formed in diploids
($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
($v=u_{31}=u_{32}$)\\
$z_i$ & trait value of individual $i$ \\
$w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$c_i$ & ploidy level of individual $i$ \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$\gamma$ & strength of directional selection \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$V$ & segregation variance in the reference diploid population \\
  $\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
$m$ & expected number of migrants per generation arriving in the new habitat \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
individual carries two copies of the same homolog, is the probability of
\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
1/6 \citep{stift2008}.
The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
formation through so-called \textit{first} and \textit{second division
restitution} \citep{storme2013}.
We summarize the expressions for the gametic segregation variance in
\cref{tbl:segvar}.
Detailed derivations can be found in \cref{s-sec:segvar}.

@@

We can simulate the mixed-ploidy infinitesimal model for a finite population
through a straightforward extension of the approach outlined in
\cite{barton2017}, provided we can efficiently track inbreeding and coancestry
coefficients across the different ploidy levels.
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$ (\cref{s-sec:tetinbred}).
The recursion for the coancestry coefficients in is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

@@

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet}).
Furthermore, we confirm that, in the absence of double reducyion, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction leads to an immediate increase in genetic
variance, but leads to accelerated inbreeding, causing faster loss of genetic
variance in the long-term (\cref{fig:vztet}B).
As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
double reduction cannot, however, be described by a single long-term effective
population size.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
\caption{
The infinitesimal model in autotetraploids.
Comparisons are shown for the decay of the genetic variance ($V_z$) due to
inbreeding in exact simulations of the infinitesimal model in autotetraploids
against individual-based simulations of autotetraploid populations with $L$
unlinked additive loci determining the quantitative trait. 
(A) Simulations of a model without double reduction ($\alpha_4=0$).
(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
all loci in the finite $L$ simulations).
We show window-smoothed values for visual clarity, with observed genetic
variances averaged in windows of 20 generations every 10 generations.
The black line marks $e^{-t/4N}$. 
We assume $N=250$ and $V_z(0) = 1$.
\label{fig:vztet}}
\end{figure}

@@

We first consider the establishment of a population from a single migrant
individual with trait value $z_0 < 0$.
We assume $u=0$ and compare the probability of establishment when the migrant
is diploid vs. tetraploid (\cref{fig:est1}).
As noted by \cite{barton2018}, the establishment probability depends
essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
$z_0/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
maladaptation, respectively.
We shall scale our results accordingly, assuming $2V=1$ throughout.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/est2.pdf}
\caption{
(B) Establishment probability from a single diploid or tetraploid individual
for increasing selection gradient $\gamma$.
(A) Probability of a tetraploid individual with trait value $z_0 = -2$
successfully founding a population ($P_4$), relative to the probability for a
diploid individual with the same trait value ($P_2$).
The vertical dashed line marks the $\beta$ value that yields equal equilibrium
variances across ploidy levels ($\beta = \sqrt{1/2}$).
All results are estimated from 500.000 replicate simulations.
\label{fig:est1}}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
\caption{
Establishment with recurrent migration.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different degrees of maladaptation
($\bar{z}_s$, the mean trait value in the source population). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
The horizontal dashed line marks the proportion of tetraploid migrants (i.e.
the proportion of tetraploids at equilibrium in the source population).
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:estwmig}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
\caption{
Establishment with recurrent migration and loss of self-incompatibility in
polyploids.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different self-fertilization rates in
polyploids. Diploids are assumed to be self-incompatible.
Note that $\sigma=0.0$ refers to random self-fertilization (i.e.
self-fertilization with probability $1/N$). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:}}
\end{figure}

>>>>

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N, N_k$ & total population size, population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new habitat \\
$w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
individual carries two copies of the same homolog, is the probability of
\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
1/6 \citep{stift2008}.
The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
formation through so-called \textit{first} and \textit{second division
restitution} \citep{bretagnolle1995,storme2013}.
We summarize the expressions for the gametic segregation variance in
\cref{tbl:segvar}.
Detailed derivations can be found in \cref{s-sec:segvar}.

@@

We can simulate the mixed-ploidy infinitesimal model for a finite population
through a straightforward extension of the approach outlined in
\cite{barton2017}, provided we can efficiently track inbreeding and coancestry
coefficients across the different ploidy levels.
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$ (\cref{s-sec:tetinbred}).
The recursion for the coancestry coefficients in is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

@@

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
Furthermore, we confirm that, in the absence of double reduction, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction leads to an immediate increase in genetic
variance, but leads to accelerated inbreeding, causing faster loss of genetic
variance in the long-term (\cref{fig:vztet,s-fig:fininf}).
As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
double reduction cannot, however, be described by a single long-term effective
population size.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
\caption{
The infinitesimal model in autotetraploids.
Comparisons are shown for the decay of the genetic variance ($V_z$) due to
inbreeding in exact simulations of the infinitesimal model in autotetraploids
against individual-based simulations of autotetraploid populations with $L$
unlinked additive loci determining the quantitative trait. 
(A) Simulations of a model without double reduction ($\alpha_4=0$).
(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
all loci in the finite $L$ simulations).
We show window-smoothed values for visual clarity, with observed genetic
variances averaged in windows of 20 generations every 10 generations.
The black line marks $e^{-t/4N}$. 
We assume $N=250$ and $V_z(0) = 1$.
\label{fig:vztet}}
\end{figure}

@@

We first consider the establishment of a population from a single migrant
individual with trait value $z_0 = 0$.
We assume $u=0$ and compare the probability of establishment when the migrant
is diploid vs. tetraploid (\cref{fig:est1}).
As noted by \cite{barton2018}, the establishment probability depends
essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
$\theta/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
maladaptation, respectively.
We shall scale our results accordingly, assuming $2V=1$ throughout.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1.pdf}
\caption{
(A) Probability of establishment from a single diploid or tetraploid individual
with trait value $z=0$ for increasing selection intensity $\gamma$.
We assume $m=0$ and $u=0$, i.e. there is no migration, and no unreduced
gametes are produced. The trait is scaled in tetraploids so as to yield the
same genetic variance at HWLE ($\beta_4 = \sqrt{1/2}$)
(B) Probability of a tetraploid individual with trait value $z = 0$
successfully founding a population ($P_4$), relative to the probability for a
diploid individual with the same trait value ($P_2$).
The vertical dashed line marks $\beta_4=\sqrt{1/2}$, for which the variance at
HWLE is identical between diploids and autotetraploids.
All results are estimated from 500.000 replicate simulations.
\label{fig:est1}}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
\caption{
Establishment with recurrent migration.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different degrees of maladaptation
($\bar{z}_s$, the mean trait value in the source population). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
The horizontal dashed line marks the proportion of tetraploid migrants (i.e.
the proportion of tetraploids at equilibrium in the source population).
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:estwmig}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
\caption{
Establishment with recurrent migration and loss of self-incompatibility in
polyploids.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different self-fertilization rates in
polyploids. Diploids are assumed to be self-incompatible.
Note that $\sigma=0.0$ refers to random self-fertilization (i.e.
self-fertilization with probability $1/N$). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:}}
\end{figure}
